---
title: "PA1_template.Rmd"
author: "Ajit Dhamale"
date: "September 18, 2016"
output: html_document
---

```{r setup}
library(data.table)
library(ggplot2)
```

# Loading and preprocessing the data
### set working directory to directory where files are downloaded
### download the input data zip file
```{r setwd}
setwd(".")
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zipName <- file.path(getwd(), "input_data.zip")
download.file(url, zipName)
unzip(zipName, overwrite = TRUE)
```

### Load the input csv in a data table and Lets look at the dimensions of input data
```{r}
csvdata <- data.table::fread("activity.csv")
dim(csvdata)
```

# What is mean total number of steps taken per day?
### now lets count steps per day by removing the NAs in the input data
```{r}
spd <- csvdata[,sum(as.numeric(steps), na.rm = TRUE), by=date]
```

# Plot histograms of steps per day
```{r}
hist(spd$V1, main = "Total number of steps per day" , xlab = "Total number of steps",breaks = 20 )
```

# report mean
```{r}
print(mean(spd$V1, na.rm =  TRUE))
```
# report median
```{r}
print(median(spd$V1, na.rm =  TRUE))
```


# What is the average daily activity pattern?

### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
### and the average number of steps taken, averaged across all days (y-axis)

### so first calculate the mean steps per interval across all days
```{r}
spi <- csvdata[,mean(as.numeric(steps), na.rm = TRUE), by=interval]
```

### now plot the average steps per interval across all days
```{r}
plot(spi$interval,spi$V1,type="l",xlab = "Interval", ylab = "Average stpes across alldays")
```

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
### Find out interval with maximum steps 
```{r}
intWithMaxSteps <- spi[which.max(spi$V1),]
print(intWithMaxSteps)
```

# Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
### Number of missing step values is 
```{r}
no_of_nas <- length((which(is.na(csvdata$steps))))
print(no_of_nas)
```

### total number of values as a percent 
```{r}
total_no_of_rows <- nrow(csvdata)
print( ( no_of_nas * 100 ) / total_no_of_rows)
```

### Strategy to fill misssing values is to fill NAs with mean of that interval across days


### tempData has original numberof steps and average number
### of steps for that day as additional column
```{r}
tempData <- merge(csvdata,spi, by.x = "interval",by.y = "interval", all.x = TRUE, sort=FALSE)
```

### rename new merged column V1 to average_steps
```{r}
names(tempData)[names(tempData) =="V1"] <- "average_steps"
```

### now add imputed_steps column to tempdata 
```{r}
tempData$imputed_steps <- NA
```

### Create a new dataset that is equal to the original dataset but with the missing data filled in
### set value of imputed steps to average stpes if steps is NA else set to steps
```{r}
tempData$imputed_steps <- ifelse(is.na(tempData$steps), tempData$average_steps, tempData$steps )
imputedData <- tempData
```

### now lets count steps per day by removing the NAs in the input data
```{r}
spdImputedData <- imputedData[,sum(as.numeric(imputed_steps), na.rm = TRUE), by=date]
```

# Make a histogram of the total number of steps taken each day using imputed data
```{r}
hist(spdImputedData$V1, main = "Total number of steps (imputed) per day" , xlab = "Total number of steps (imputed)",breaks = 20 )
```


# report mean for imputed data
```{r}
print(mean(spdImputedData$V1, na.rm =  TRUE))
```

# report median for imputed data
```{r}
print(median(spdImputedData$V1, na.rm =  TRUE))
```

### note that mean and median has increased after filling in the missing values


# Are there differences in activity patterns between weekdays and weekends?
### add day column based on date
```{r}
imputedData$day <- weekdays(as.Date(imputedData$date), abbreviate =  TRUE)
```

### add factor column to identify weekdays
```{r}
imputedData$is_weekday <- as.factor(!( imputedData$day  == "Sun" | imputedData$day == "Sat"))
```

### Calculate steps per interval using imputed data
```{r}
spiImputedData <- imputedData[,mean(as.numeric(steps), na.rm = TRUE), by=interval]
```

# Create a new factor variable in the dataset with two levels - "weekday" and "weekend" 
### Now calculate the average number of steps per interval for WEEKDAYS
```{r}
cw <- imputedData[which(imputedData$is_weekday == TRUE ),mean(as.numeric(steps), na.rm = TRUE), by=interval]
```
### For the results assign factor 'weekday'
```{r}
cw$day_type <- as.factor("Weekday")
```

### Now calculate the average number of steps per interval for WEEKENDS days
```{r}
cwknd <- imputedData[which(imputedData$is_weekday == FALSE ),mean(as.numeric(steps), na.rm = TRUE), by=interval]
```

### For the results assign factor 'Weekend'
```{r}
cwknd$day_type <- as.factor("Weekend")
```

### finally combine the average steps per interval for weekdays and weekends in a single data frame
```{r}
finalData <- rbindlist(list(cw,cwknd))
```

### Now plot the results for Average number of steps per interval, add facets for day_type
```{r}
finalPlot <- qplot(interval,V1,data=finalData,geom="line",ylab =  "Average number of stpes", xlab = "Interval")
finalPlot <- finalPlot + facet_wrap( ~ day_type, ncol = 1)
print(finalPlot)
```

